A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-18 tex/body.tex
uniformized style

\begin{lemma}
\label{lem:interpretation-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\ValC$ is a completion of $\Val$ for a normal interpretation $\J$, the following properties hold:
\begin{enumerate}
\item[\rm (1)] $\ValC \ec{x\typ{\tau}}$ is $\alpha$-equivalent to a value in $\J( \tau )$.
\item[\rm (2)]
$\ValC \ec{x} = \interp{\ValC \ec{y}}{ t }$ for
all subterms $t$ of $\ValC \ec{y}$ with $\muvar(t) = {\Varec{x}}$.
\item[\rm (3)]
$\ValC \ec{x} \vsim \ValC \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{rep}
\begin{proof}
To show 1, we first show that $\ValC$ contains no free variables.
Assume by contradiction that $\ValC$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be $1$.
Since \rn{Single} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$ of type $\tau$.
This is a contradiction, since %our %model
completion can assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$,
and thus %our %model
completion can assign a value not occurring in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$ to ${\Varec{y}}$.
By construction, $\ValC \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\ValC$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\ValC \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\ValC \ec{x}$ is $\alpha$-equivalent to a value in $\J( \tau )$.

We first show that 2 and 3 hold initially for $\Val$.
For all equivalence classes $\ec{z}$,
each pair of constructor terms $\const{C}_{\negvthinspace j}( \vec t )$ and $\const{C}_{\negvthinspace\jPrime}( \vec u )$ in $\ec{z}$
are such that $j = \jPrime$ since \rn{Clash} does not apply to $\Ec$,
and are such that $\ec{ \vec t } = \ec{ \vec u }$ since \rn{Inject} does not apply to $\Ec$.
Thus,
$\Val$ was constructed by applying a sequence of substitutions
where all substitutions for variables ${\Varec{z}}$
were uniquely of the form $\{ {\Varec{z}} \mapsto \const{C}_{\negvthinspace j}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}$
when $\ec{z}$ contains a constructor $\const{C}_{\negvthinspace j}( t_1, \ldots, t_n )$.
Say $\Val \ec{y}$ has a subterm $t$ where $\muvar(t) = {\Varec{x}}$.
Both $\Val \ec{x}$ and the subterm $t$ of $\Val \ec{y}$ were
constructed by applying a sequence of substitutions of the form mentioned above to ${\Varec{x}}$.
Moreover, free variables ${\Varec{z}}$ in $t$ that are bound in $\Val \ec{y}$ are interpreted in
the expansion of $\interp{\Val \ec{y}}{ t }$ as a term
constructed by a sequence of substitutions of the form mentioned above to ${\Varec{z}}$. %\rem{TODO: more detail}
Thus, we have
$\interp{\Val \ec{y}}{ t } = \interp{\Val \ec{x}}{ \Val \ec{x} } = \Val \ec{x}$, and thus 2 holds for $\Val$.
Statement 3 holds for $\Val$ since \rn{Unique} does not apply.

@@

We now show that if (2) and (3) hold for some $\Val_1$,
they also hold for $\nf{\Val_1 \sigma}$,
where $\sigma$ is a substitution of the form $\{ {\Varec{x}} \mapsto \MU{\Varec{x}}.\; t \}$,
${\Varec{x}} \in FV( \Val_1 )$,
and $\MU {\Varec{x}}.\; t$ is not $\alpha$-equivalent to a term in $\Terms^{\smash{\J}}_{\Varec{x}}(\Val_1)$.
To show 2,
by assumption of 2 on $\Val_1$, we have $\interp{\Val_1 \ec{y}}{ u } = \Val_1 \ec{x}$ for
all subterms $u$ of $\Val_1 \ec{y}$ where $\muvar(t) = {\Varec{x}}$.
Thus, $\interp{\Val_1 \sigma \ec{y}}{ u } = \Val_1 \sigma \ec{x}$ and
$\interp{\nf{\Val_1 \sigma} \ec{y}}{ u } = \nf{\Val_1 \sigma} \ec{x}$.
To show 3,
consider two distinct equivalence classes $\ec{y}$ and $\ec{z}$,
and assume by contradiction that $\nf{\Val_1 \sigma} \ec{y} \aequiv \nf{\Val_1 \sigma} \ec{z}$.
Due to (3) for $\Val_1$,
$\ec{y}$ and $\ec{z}$ must have (minimal) subterms
where $t_1$ occurs in $\Val_1 \ec{y}$ the same position (call it $p$) as $t_2$ occurs in $\Val_1 \ec{z}$,
and $t_1 \not\aequiv t_2$.
%We first show that $\nf{\Val_1 \sigma} \ec{y}$ (and for the same reasons, $\nf{\Val_1 \sigma} \ec{z}$)
%also contains position $p$.
%If this were not the case,
%then without loss of generality we may assume
%$\Val_1 \sigma \ec{y}$ contains a self-similar subterm
%$\MU {\Varec{w}}.\; u$ that strictly contains $t_1 \sigma$.
If $t_1$ (resp. $t_2$) is a free variable that is not ${\Varec{x}}$,
then $\nf{\Val_1 \sigma} \ec{y}$ (resp. $\nf{\Val_1 \sigma} \ec{z}$)
contains $t_1$ (resp. $t_2$) at position $p$,
and $\nf{\Val_1 \sigma} \ec{z}$ (resp. $\nf{\Val_1 \sigma} \ec{y}$) does not.
If $t_1$ is of the form $\MU w_1.\; \const{C}( \vec t )$
and $t_2$ is of the form $\MU w_2.\; \const{D}( \vec u )$,
then the expansion %\rem{TODO: define}
of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$.
Since $t_1$ and $t_2$ are minimal,
say $t_1$ is of the form $\MU w.\; \const{C}( \vec t )$,
and $t_2$ is ${\Varec{x}}$.
Since $\sigma$ maps ${\Varec{x}}$ to a closed $\mu$-term $\MU {\Varec{x}}.\; t$,
we have that $FV( t_1 ) \subseteq \{ {\Varec{x}} \}$,
or else the expansion of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$ since they do not contain the same free variables.
Since $\interp{\Val_1 \ec{y}}{t_1}$ is normal,
there is a closed $\mu$-term $v \in \Terms^{\smash{\J}}_{\Varec{x}}( \Val_1 )$
such that $v \vsimv{\Varec{x}} \interp{\Val_1 \ec{y}}{t_1}$.
Thus, by assumption on our selection of $\MU {\Varec{x}}.\; t$,
we have $\MU {\Varec{x}}.\; t \not\vsimv{\Varec{x}} \interp{\Val_1 \ec{y}}{t_1}$,
which implies that the expansion of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$. %\rem{TODO: more detail}

>>>>

\begin{lemma}
\label{lem:interpretation-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\ValC$ is a completion of $\Val$ for a normal interpretation $\J$, the following properties hold:
\begin{enumerate}
\item[\rm (1)] $\ValC \ec{x\typ{\tau}}$ is $\alpha$-equivalent to a value in $\J( \tau )$.
\item[\rm (2)]
$\ValC \ec{x} = \interp{\ValC \ec{y}}{ t }$ for
all subterms $t$ of $\ValC \ec{y}$ with $\muvar(t) = {\Varec{x}}$.
\item[\rm (3)]
$\ValC \ec{x} \vsim \ValC \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{rep}
\begin{proof}
To show (1), we first show that $\ValC$ contains no free variables.
Assume by contradiction that $\ValC$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Single} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$ of type $\tau$.
This is a contradiction, since %our %model
completion can assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$,
and thus %our %model
completion can assign a value not occurring in $\Terms^{\smash{\J}}_{\Varec{y}}(\ValC)$ to ${\Varec{y}}$.
By construction, $\ValC \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\ValC$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\ValC \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\ValC \ec{x}$ is $\alpha$-equivalent to a value in $\J( \tau )$.

We first show that (2) and (3) hold initially for $\Val$.
For all equivalence classes $\ec{z}$,
each pair of constructor terms $\const{C}_{\negvthinspace j}( \vec t )$ and $\const{C}_{\negvthinspace\jPrime}( \vec u )$ in $\ec{z}$
are such that $j = \jPrime$ since \rn{Clash} does not apply to $\Ec$,
and are such that $\ec{ \vec t } = \ec{ \vec u }$ since \rn{Inject} does not apply to $\Ec$.
Thus,
$\Val$ was constructed by applying a sequence of substitutions
where all substitutions for variables ${\Varec{z}}$
were uniquely of the form $\{ {\Varec{z}} \mapsto \const{C}_{\negvthinspace j}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}$
when $\ec{z}$ contains a constructor $\const{C}_{\negvthinspace j}( t_1, \ldots, t_n )$.
Say $\Val \ec{y}$ has a subterm $t$ where $\muvar(t) = {\Varec{x}}$.
Both $\Val \ec{x}$ and the subterm $t$ of $\Val \ec{y}$ were
constructed by applying a sequence of substitutions of the form mentioned above to ${\Varec{x}}$.
Moreover, free variables ${\Varec{z}}$ in $t$ that are bound in $\Val \ec{y}$ are interpreted in
the expansion of $\interp{\Val \ec{y}}{ t }$ as a term
constructed by a sequence of substitutions of the form mentioned above to ${\Varec{z}}$. %\rem{TODO: more detail}
Thus, we have
$\interp{\Val \ec{y}}{ t } = \interp{\Val \ec{x}}{ \Val \ec{x} } = \Val \ec{x}$, and thus (2) holds for $\Val$.
Property (3) holds for $\Val$ since \rn{Unique} does not apply.

@@

We now show that if (2) and (3) hold for some $\Val_1$,
they also hold for $\nf{\Val_1 \sigma}$,
where $\sigma$ is a substitution of the form $\{ {\Varec{x}} \mapsto \MU{\Varec{x}}.\; t \}$,
${\Varec{x}} \in FV( \Val_1 )$,
and $\MU {\Varec{x}}.\; t$ is not $\alpha$-equivalent to a term in $\Terms^{\smash{\J}}_{\Varec{x}}(\Val_1)$.
To show (2),
by assumption of (2) on $\Val_1$, we have $\interp{\Val_1 \ec{y}}{ u } = \Val_1 \ec{x}$ for
all subterms $u$ of $\Val_1 \ec{y}$ where $\muvar(t) = {\Varec{x}}$.
Thus, $\interp{\Val_1 \sigma \ec{y}}{ u } = \Val_1 \sigma \ec{x}$ and
$\interp{\nf{\Val_1 \sigma} \ec{y}}{ u } = \nf{\Val_1 \sigma} \ec{x}$.
To show (3),
consider two distinct equivalence classes $\ec{y}$ and $\ec{z}$,
and assume by contradiction that $\nf{\Val_1 \sigma} \ec{y} \aequiv \nf{\Val_1 \sigma} \ec{z}$.
Due to (3) for $\Val_1$,
$\ec{y}$ and $\ec{z}$ must have (minimal) subterms
where $t_1$ occurs in $\Val_1 \ec{y}$ the same position (call it $p$) as $t_2$ occurs in $\Val_1 \ec{z}$,
and $t_1 \not\aequiv t_2$.
%We first show that $\nf{\Val_1 \sigma} \ec{y}$ (and for the same reasons, $\nf{\Val_1 \sigma} \ec{z}$)
%also contains position $p$.
%If this were not the case,
%then without loss of generality we may assume
%$\Val_1 \sigma \ec{y}$ contains a self-similar subterm
%$\MU {\Varec{w}}.\; u$ that strictly contains $t_1 \sigma$.
If $t_1$ (resp. $t_2$) is a free variable that is not ${\Varec{x}}$,
then $\nf{\Val_1 \sigma} \ec{y}$ (resp. $\nf{\Val_1 \sigma} \ec{z}$)
contains $t_1$ (resp. $t_2$) at position $p$,
and $\nf{\Val_1 \sigma} \ec{z}$ (resp. $\nf{\Val_1 \sigma} \ec{y}$) does not.
If $t_1$ is of the form $\MU w_1.\; \const{C}( \vec t )$
and $t_2$ is of the form $\MU w_2.\; \const{D}( \vec u )$,
then the expansion %\rem{TODO: define}
of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$.
Since $t_1$ and $t_2$ are minimal,
say $t_1$ is of the form $\MU w.\; \const{C}( \vec t )$,
and $t_2$ is ${\Varec{x}}$.
Since $\sigma$ maps ${\Varec{x}}$ to a closed $\mu$-term $\MU {\Varec{x}}.\; t$,
we have that $FV( t_1 ) \subseteq \{ {\Varec{x}} \}$,
or else the expansion of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$ since they do not contain the same free variables.
Since $\interp{\Val_1 \ec{y}}{t_1}$ is normal,
there is a closed $\mu$-term $v \in \Terms^{\smash{\J}}_{\Varec{x}}( \Val_1 )$
such that $v \vsimv{\Varec{x}} \interp{\Val_1 \ec{y}}{t_1}$.
Thus, by assumption on our selection of $\MU {\Varec{x}}.\; t$,
we have $\MU {\Varec{x}}.\; t \not\vsimv{\Varec{x}} \interp{\Val_1 \ec{y}}{t_1}$,
which implies that the expansion of $\nf{\Val_1 \sigma} \ec{y}$ and $\nf{\Val_1 \sigma} \ec{z}$
are $\alpha$-disequivalent at position $p$. %\rem{TODO: more detail}
